A Tutorial on Tailored Simulation-Based Power Analysis for Experimental Designs with Generalized Linear Mixed Models

### 2023-09-18 docs/manuscript.tex
updated repo for preprint publication

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{power analysis, data simulation, sample size, generalized linear mixed model\newline\indent Word count: 7059}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={A Tutorial on Tailored Power Simulations for Experimental Study Designs with Generalized Linear Mixed Models},
  pdfauthor={Timo K. Koch,1, 2, Florian Pargent,1, Anne-Kathrin Kleine1, Eva Lermer1,3, \& Susanne Gaube1,4},
  pdflang={en-EN},
  pdfkeywords={power analysis, data simulation, sample size, generalized linear mixed model},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{A Tutorial on Tailored Power Simulations for Experimental Study Designs with Generalized Linear Mixed Models}
\author{Timo K. Koch\textsuperscript{$\dagger{}$,1, 2}, Florian Pargent\textsuperscript{$\dagger{}$,1}, Anne-Kathrin Kleine\textsuperscript{1}, Eva Lermer\textsuperscript{1,3}, \& Susanne Gaube\textsuperscript{1,4}}
\date{}

\shorttitle{GLMM Power Simulation Tutorial}

@@

@@

Correspondence concerning this article should be addressed to Timo K. Koch. E-mail: \href{mailto:timo.koch@unisg.ch}{\nolinkurl{timo.koch@unisg.ch}}

@@

\abstract{%
{[}THIS WILL BE UPDATED BASED ON GOOGLE DOC{]} Researchers often lack the tools to conduct a priori power analysis for complex experimental designs deployed in human-computer interaction (HCI) studies. Generalized Linear Mixed Models (GLMMs) are well-suited to analyze experiments in which subjects respond to a set of stimuli. However, existing software packages can only be used to simulate power for simple designs, which are often not useful for more sophisticated investigations. This tutorial addresses this gap by providing a comprehensive tutorial to perform tailored power analyses using data simulation. Our tutorial provides code for an HCI case study, empowering researchers to simulate their own data and ascertain appropriate sample sizes for GLMMs. This resource should help to enhance the precision of study designs and to mitigate the risk of underpowered studies. Moreover, we discuss further implications for experimental research and provide an outlook on the increasing relevance of power simulations in the field.
}

@@

When planning experimental research, it is essential to determine an appropriate sample size (Lakens, 2022b) to ensure that the results obtained are both robust and informative, and to use appropriate statistical models to analyze the data. However, the recent replication crisis in experimental research, such as in Human-Computer Interaction (HCI), has illustrated many challenges surrounding the reproducibility and reliability of findings (Robertson \& Kaptein, 2016). As a result, there is a growing need for rigorous statistical methodology and the adoption of well-powered experimental designs. While software solutions exist for simple statistical models and experimental designs, many researchers lack the skills and tools to conduct power analyses for more complex research designs using the flexible generalized linear mixed models (GLMM) framework. In the present work, we provide a tutorial comprised of a concrete example for custom a priori power analyses using data simulations based on GLMMs.

@@

In empirical research relying on hypothesis testing, the most common strategy for determining an adequate sample size is based on statistical power (Lakens, 2022b). Statistical power is defined as the probability that a hypothesis test has a significant p-value when analyzing repeated samples from a population with a true effect of some pre-specified size. Less formally, power is described as the probability that a hypothesis test accepts the alternative hypothesis \(H_1\) if \(H_1\) is indeed true. If the sample size (i.e., the number of participants and/or stimuli) used for data collection is insufficient to detect the effects with high probability the effects or relationships being investigated , this particular study would be considered ``underpowered''.

Conducting underpowered research has many negative consequences. First, relying on underpowered experiments may yield inconclusive (if low power is taken into account) or misleading (if low power is ignored) results, hindering the accumulation of knowledge. Second, underpowered studies waste resources by consuming time, effort, and funding without delivering meaningful results.

@@

A power analysis represents the act of calculating the statistical power for a given true effect and sample size.
By running the calculation for a range of sample sizes ``a priori'' (i.e., before data collection), the required sample size can be determined so that researchers find an assumed true effect with the desired statistical power.
Thereby, power analysis offers a valuable contribution to the research process by allowing researchers to estimate the appropriate sample sizes required to achieve sufficient statistical power for results with high evidential value.

Conducting a careful a priori power analysis helps researchers decide which experimental design and statistical models are both feasible and appropriate for analyzing the data and answering their research questions.
Also, when conducting a proper power analysis, researchers have to consider every aspect of the experimental design and will notice statistical or design challenges before starting with data collection.
Especially in fast-moving research fields, adding a solid sample size calculation to the process can act as a safeguard for ensuring high-quality research.

Moreover, many journals and funding agencies now require that a power analysis is included in study protocols and grant proposals, recognizing its significance in ensuring robust and meaningful findings. By incorporating power analysis into research planning, researchers can enhance the replicability and credibility of their work, contributing to the advancement of the respective research field through more robust insights.

For simple statistical models, like t-tests, ANOVA, and linear regression, with common study designs (e.g., mean comparison between two groups), user-friendly software for power analysis is readily available (Faul, Erdfelder, Buchner, \& Lang, 2009).

\hypertarget{generalized-linear-mixed-models-glmm}{%
\subsection{Generalized linear mixed models (GLMM)}\label{generalized-linear-mixed-models-glmm}}

As study designs become more complex, researchers require more sophisticated statistical models to capture the nuanced relationships and hierarchical structures introduced by their study designs (Yarkoni, 2022). GLMMs (also called multilevel models) are gaining increasing popularity in analyzing data from experimental research, such as in HCI, because they offer a flexible framework for analyzing data with non-normal and categorical outcomes while accounting for both fixed and random effects (Fahrmeir, Kneib, Lang, \& Marx, 2021; Kaptein, 2016).

GLMMs are an extension of LMMs (Linear Mixed Models), which are, in turn, extensions of classic linear regression models that account for correlated data and hierarchical structures (Fahrmeir et al., 2021). In this context, correlated data means that observations within a given dataset are not independent of each other. In other words, the values of the dependent variable for one observation may be related to the values of the dependent variable for other observations in a systematic way. This correlation can arise for various reasons, such as repeated measurements on the same subjects over time, observations clustered within certain groups or locations, or data collected at different levels of granularity. Hierarchical structures can be found in datasets that have a nested or hierarchical organization. For example, one might have data on students nested within schools, employees nested within departments, or patients nested within hospitals. In such cases, the observations within each higher-level grouping (e.g., school, department, hospital) tend to be more similar to each other than to observations in other groups.
LMMs are used when the outcome variable is continuous and follows a normal distribution (when conditioned on the predictor variables). They allow for the modelling of fixed effects, which capture the relationships between predictors and the outcome, as well as random effects, which account for the correlation and variability within groups or subjects. Random effects are typically assumed to follow a normal distribution with a mean of zero and a variance that quantifies the heterogeneity across the groups or subjects.
As mentioned, GLMMs extend the LMM framework to accommodate non-normal and categorical outcome variables. They are used when the outcome variable does not follow a normal distribution but instead belongs to a different distribution family (e.g., binomial) . GLMMs incorporate both fixed and random effects, similar to LMMs, but also involve a link function that connects the linear predictor to the expected value of the outcome variable. The link function allows for modelling the relationship between predictors and the outcome in a non-linear way that is appropriate for the specific distribution family of the response variable.

\hypertarget{power-simulations-for-glmms}{%
\subsection{Power simulations for GLMMs}\label{power-simulations-for-glmms}}

Currently available software packages for power simulations do not include GLMMs, making it necessary to build data simulations tailored specifically to the study design (Murayama, Usami, \& Sakaki, 2022; Westfall, Kenny, \& Judd, 2014).
A number of tutorials have been published describing how to perform such power simulation for multilevel models (Arend \& Schäfer, 2019; Brysbaert \& Stevens, 2018; DeBruine \& Barr, 2021; Kumle, Võ, \& Draschkow, 2021; Lafit et al., 2021; Zimmer, Henninger, \& Debelak, 2022).
However, most of these tutorials only cover linear mixed models (LMMs) and focus on the most common designs (e.g., INSERT EXAMPLE). This narrow focus provides little guidance for researchers faced with more complex study designs (e.g., INSERT EXAMPLE), which are very common (Brown, 2021).
The necessary assumptions for power simulations in GLMMs include assumptions about the distributional form of the outcome variable, the random effects, and the error structure. The distributional assumption specifies the family of distributions for the outcome variable, such as Gaussian, Poisson, or binomial. Assumptions about the random effects include the assumption of normality and the covariance structure among the random effects. Additionally, assumptions about the error structure, such as independence or correlation, must be specified. Interpreting these assumptions entails understanding the underlying assumptions of the model and ensuring they align with the characteristics of the data being analyzed.
Existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, specifying assumptions a practical challenge (Maxwell, Kelley, \& Rausch, 2008).

Based on the need for well-powered experimental research using GLMMs and the lack of tools to conduct corresponding power analyses, in this tutorial paper, we present a case study that serves as a practical demonstration of how to perform a data simulation and an a priori power simulation for a more complex study design. Thereby, we aim to equip researchers with the tools needed to simulate their own data and determine appropriate sample sizes for their own research.

@@

In this section, we outline the steps for performing data simulation and a priori power analysis for GLMMs using a case study based on a specific experimental study design from the area of human-AI (artificial intelligence) interaction research.

@@

In the present case study, we simulate the data for an experiment where the diagnostic performance of users of an AI-enabled diagnostic decision support system is to be evaluated.
The goal is to understand how AI advice influences medical decision-making. Participants, radiologists (task experts) and students/interns (non-task experts), review head computer tomography (CT) scans to assess the presence of bleeding, more specifically, intracranial haemorrhage (ICH).
To support their decision-making, in two experimental conditions , an AI model provides initial diagnostic advice, which can be used as guidance by the participants. This AI advice can be either correct (80\% of cases) or incorrect (20\%). In the control condition, no AI advice is presented, meaning that the participants have to read the CT scan without any support. After reviewing the CT scan, participants deliver a medical diagnosis (bleeding or no bleeding), which may be either accurate or inaccurate.
This experimental design introduces some missing values by design since the advice is neither correct nor incorrect when no advice is present, which must be taken into account when simulating and analyzing the data. With this experiment, we want to determine if (a) experts are better than non-experts in reading head CT scans, (b) correct AI advice leads to better diagnostic accuracy than incorrect AI advice, and (c) different presentations of AI advice influence performance.
In this example, recruiting task experts (i.e., radiologists) is more challenging due to their limited availability, while non-experts (i.e., students/interns) are more readily accessible.
The goal of the power simulation is to determine how many task experts and non-experts must be recruited to achieve sufficient statistical power in the planned experiment.

@@

In our case study, we use the lme4 R package (Bates, Mächler, Bolker, \& Walker, 2015) which is a state-of-the-art tool for fitting frequentist GLMMs.\footnote{For Bayesian GLMMs, the brms R package is currently the most prominent option (Bürkner, 2017).}

The lme4 package includes a function called \texttt{simulate} that allows researchers to simulate the dependent variable based on the same model formula used for model fitting, enabling power simulations and other related analyses.

However, the model parameterization used by the lme4 package is quite technical, making it difficult for applied researchers to determine whether their specified population model (i.e., the theoretical model that describes the underlying data generation process for a specific population of interest) implies plausible associations in their simulated data. Therefore in this tutorial, we simulate data for GLMMs from first principles (i.e., creating synthetic data step by step instead of using black box functions) to assist applied researchers in better understanding all model assumptions and then use lme4 to analyze the simulated data sets.\footnote{A less flexible alternative would be to use the simr package (Green \& MacLeod, 2016), which can be used to both simulate data and perform power analysis for models supported by the lme4 package.}

@@

In a GLMM, the expected value of the dependent variable Y conditioned on the vector of predictor variables \(\mathbf{X}\) and random effects \(\mathbf{U}\), transformed by a link function \(g()\) is modeled as a linear combination \(\eta\) of the predictor variables \(\mathbf{X}\), the random effects \(\mathbf{U}\) and the model parameters \(\mathbf{\beta}\) (Fahrmeir et al., 2021). \[
g(E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = \eta <!-- SG: Sorry for maybe a stupid question, but what does E stand for? and why are model parameters beta mentioned but not in the equation?-->
\] Equivalently, the conditional expected value is modelled as the linear combination \(\eta\), transformed by the inverse link function \(g^{-1}()\). \[
E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = g^{-1}(\eta)
\] If the dependent variable (i.e., diagnostic decision) \(Y\) is a binary variable with values \(0\) (i.e.~inaccurate), or \(1\) (i.e., accurate), the conditional expected value is equivalent to the probability: \[
P_{si} := P(Y = 1|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})
\] In our case study, \(P_{si}\) is the conditional probability that a subject \(s\) gives the correct response to item (i.e., CT scan) \(i\).

@@

In our case study, the probability of giving a correct response is assumed to depend on the predictors:

\begin{itemize}
\tightlist
\item
  \(advice\_present_{si}\): whether subject \(s\) was presented with AI advice (1) or not (0) when asked to asses item \(i\)
\item
  \(advice\_correct_{si}\): whether this advice was correct (1) or not (0)
\item
  \(expert_s\): whether subject \(s\) was a task expert (1) or not (0)
\end{itemize}

@@

In the GLMM literature, this would be called a binomial GLMM with two random intercepts (for subjects and items), two level-1 predictors (\(advice\_present\), \(advice\_correct\)), one level-2 predictor (\(expert\)) and two cross-level interactions (\(expert \cdot advice\_present\), \(expert \cdot advice\_correct\)).

To limit complexity, we do not consider random slopes, additional predictors or higher-level interactions here.

@@

In the first six lines of the function definition, we set some default parameter values (which we will explain in a later section) and load the packages we use to manipulate and simulate data.
In our case study, each subject (\texttt{n\_subjects} in total) is assumed to respond to each item (i.e., CT scan; \texttt{n\_items} in total). Thus, the \texttt{add\_random} command creates a fully-crossed \texttt{data.frame} with \texttt{n\_subjects} \(\times\) \texttt{n\_items} rows. We add a between-subject effect with the \texttt{add\_between} command, simulating that about \(25\%\) of subjects are experts. The next two lines simulate that in \(\frac{2}{3}\) of trials, subjects will be presented with AI advice, and if advice is presented, the advice will be correct in about \(80\%\) of cases (the variable \texttt{advice\_correct} is always 0 when no advice is presented). Next, we simulate one random effect for each subject (\texttt{u0s}) and for each item (\texttt{u0i}). As assumed by standard GLMMs, the \texttt{add\_ranef} function draws the random effects from a normal distribution with a mean 0 and a standard deviation specified by the user. With all design variables done, we are ready to simulate our model equation as outlined in equation X. The linear predictor variable \texttt{linpred} (\(\eta\) in the GLMM model equations) combines the predictor variables, random effects and model parameters as assumed by our model. We then transform the linear predictor with the inverse-link function to compute \texttt{y\_prob}, the probability that the subject correctly solved the item (in R, the inverse-logit link is computed with \texttt{plogis} and the logit link with \texttt{qlogis}). In the final step, we simulate the binary dependent variable \texttt{y\_bin} (i.e., whether the subject chooses the accurate diagnosis for the CT scan) by -- for each trial -- drawing from a Bernoulli distribution with success probability \texttt{y\_prob}.

@@

In this section, we show how to fit a GLMM with lme4, interpret the model and test hypotheses derived from a research question.

We simulate data according to our model, in which 100 subjects respond to 50 items (we use \texttt{set.seed} to make the simulation reproducible). However, for the sake of the exercise, we can imagine that this would be real data resulting from our future experiment and think about how we would analyse this data.

@@

@@

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{power analysis, data simulation, sample size, generalized linear mixed model\newline\indent Word count: 7598}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={A Tutorial on Tailored Simulation-Based Power Analysis for Experimental Designs with Generalized Linear Mixed Models},
  pdfauthor={Timo K. Koch,1, 2, Florian Pargent,1, Anne-Kathrin Kleine1, Eva Lermer1,3, \& Susanne Gaube1,4},
  pdflang={en-EN},
  pdfkeywords={power analysis, data simulation, sample size, generalized linear mixed model},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{A Tutorial on Tailored Simulation-Based Power Analysis for Experimental Designs with Generalized Linear Mixed Models}
\author{Timo K. Koch\textsuperscript{$\dagger{}$,1, 2}, Florian Pargent\textsuperscript{$\dagger{}$,1}, Anne-Kathrin Kleine\textsuperscript{1}, Eva Lermer\textsuperscript{1,3}, \& Susanne Gaube\textsuperscript{1,4}}
\date{}

\shorttitle{Simulation-Based Power Analysis for GLMMs}

@@

This research was funded by a grant from the Volkswagen Foundation (Grant No.~98525).
All R code in this manuscript is available in the project's repository on the Open Science Framework (OSF): \url{https://osf.io/dhwf4/}

@@

Correspondence concerning this article should be addressed to Timo K. Koch, Torstrasse 25, 9000 St.~Gallen, Switzerland. E-mail: \href{mailto:timo.koch@unisg.ch}{\nolinkurl{timo.koch@unisg.ch}}

@@

\note{\clearpage}

\abstract{%
When planning experimental research, determining an appropriate sample size and using suitable statistical models are crucial for robust and informative results. However, the recent replication crisis in Human-Computer Interaction (HCI) and other empirical research fields underlines the need for more rigorous statistical methodology and well-powered designs. Generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. Yet, analytic methods and software cannot be applied to conduct a priori power analyses for GLMMs, necessitating data simulation approaches. Based on a practical case study, the current tutorial equips researchers with a step-by-step guide and corresponding code for conducting tailored a priori power analyses to determine appropriate sample sizes with GLMMs. Finally, we give an outlook on the increasing importance of simulation-based power analysis in experimental research.
}

@@

When planning experimental research, it is essential to determine an appropriate sample size to ensure that the results obtained are both robust and informative, and to use appropriate statistical models to analyze the data (Lakens, 2022b). However, the recent replication crisis in Human-Computer Interaction (HCI) and several other disciplines grounded on empirical research has illustrated many challenges surrounding the reproducibility and reliability of findings (Cockburn, Dragicevic, Besançon, \& Gutwin, 2020; Robertson \& Kaptein, 2016; Yarkoni, 2022). As a result, there is a growing need for more rigorous statistical methodology and the adoption of well-powered experimental designs. While software solutions exist for simple statistical models and experimental designs, many researchers lack the skills and tools to conduct ``a priori'' (i.e., before data collection) power analyses for more complex research designs using the flexible generalized linear mixed models (GLMM) framework in order to determine the required sample size in their experiments. In the present work, we provide a tutorial consisting of a concrete example for tailored a priori power analyses using data simulations based on GLMMs.

@@

In empirical research relying on hypothesis testing, the most common strategy for determining an adequate sample size is based on statistical power (Lakens, 2022b). Statistical power is defined as the probability that a hypothesis test has a significant p-value when analyzing repeated samples from a population with a true effect of some pre-specified size. Less formally, power is described as the probability that a hypothesis test correctly rejects the null hypothesis when the alternative hypothesis is true. If the sample size (i.e., the number of participants and/or stimuli) used for data collection is insufficient to detect the effects or relationships being investigated with high probability, the study would be considered ``underpowered''.

Conducting underpowered research has many negative consequences. First, relying on underpowered experiments may yield inconclusive (if researchers acknowledge the small evidential value of an underpowered study in the limitations section) or misleading (if low power is ignored by the researchers) results, hindering the accumulation of knowledge. Second, underpowered studies waste resources by consuming time, effort, and funding without delivering meaningful results.

@@

A power analysis represents the act of calculating the statistical power for a given true effect and sample size.
When running a power analysis before data collection, the required sample size can be determined so that researchers find an assumed true effect with the desired statistical power.

Thereby, a priori power analysis offers a valuable contribution to the research process by allowing researchers to estimate the appropriate sample sizes required to achieve sufficient statistical power for results with high evidential value.
Moreover, conducting a careful a priori power analysis helps researchers decide which experimental design and statistical models are both feasible and appropriate for analyzing the data and answering their research questions.
Also, when conducting a proper power analysis, researchers have to consider every aspect of the experimental design and will notice statistical or design challenges before starting with data collection.
Adding a solid sample size calculation to the research process can act as a safeguard for ensuring high-quality research.
Finally, many journals and funding agencies now require that a power analysis is included in study protocols and grant proposals, recognizing its significance in ensuring robust and meaningful findings.

For simple statistical models, like t-tests, ANOVA, and linear regression, with common study designs (e.g., mean comparison between two groups), user-friendly software for a priori power analysis is readily available (Champely et al., 2018; Faul, Erdfelder, Buchner, \& Lang, 2009).
However, these software packages are often not flexible enough to perform power analysis for complex designs.

\hypertarget{generalized-linear-mixed-models-glmms}{%
\subsection{Generalized linear mixed models (GLMMs)}\label{generalized-linear-mixed-models-glmms}}

As study designs become more complex, researchers require more sophisticated statistical models to capture the nuanced relationships and grouping structures introduced by their study designs (Yarkoni, 2022). GLMMs (also called multilevel models) are gaining increasing popularity in analyzing data in HCI and other empirical disciplines because they offer a flexible framework for handling data with outcome variables that are not normally distributed (e.g., categorical outcomes) while accounting for both fixed and random effects (Fahrmeir, Kneib, Lang, \& Marx, 2021; Kaptein, 2016).

GLMMs are an extension of LMMs (Linear Mixed Models), which are, in turn, extensions of linear regression models that account for correlated data including hierarchical structures (Fahrmeir et al., 2021).
In this context, correlated data means that the value in the outcome variable for one observation may be related (i.e., more similar or less similar) to the value for another observation in a systematic way that is not already accounted for by the usual (fixed) predictor variables (e.g., age of participants). This correlation can arise for various reasons: Responses to some stimuli from some participants might be more similar because the same person was measured twice (repeated measurements), both participants come from the same neighborhood (clustering) or both participants responded to the same stimulus (stimulus effects).
Thus, modeling such correlations is especially important whenever the data has a clear structure, while the grouping variables can be hierarchically organized (e.g., students nested in schools, schools nested in districts) or not (e.g., students solve math exercises, but neither student sees all exercises).
LMMs are used when the outcome variable is continuous and follows a normal distribution (when conditioned on all fixed and random effects).
They allow for the modeling of fixed effects, which capture the relationships between our usual predictors and the outcome, as well as random effects, which account for the different types of correlation structure and grouping effects exemplified above.
Random effects are typically assumed to follow a normal distribution with a mean of zero and a variance that quantifies the heterogeneity across groups.

As mentioned, GLMMs extend the LMM framework to accommodate non-normally distributed continuous and categorical outcome variables.
GLMMs incorporate both fixed and random effects, similar to LMMs, but also involve a link function that connects the linear combination of predictor variables to the expected value of the outcome variable. The link function allows for modeling the relationship between predictors and the outcome in a non-linear way that is appropriate for the specific distribution family of the outcome variable.
As an example, think of an experiment with different design factors (e.g., picture, headline) impacting the likelihood of users clicking on an online advertisement. Here, participants' behavior is measured repeatedly (e.g., over several sessions). The click patterns of participants in one session are likely to be correlated with their previous sessions. Finally, the outcome variable is binary (click/no click) for each interaction, which follows a binomial distribution.

\hypertarget{power-analysis-for-glmms}{%
\subsection{Power analysis for GLMMs}\label{power-analysis-for-glmms}}

Power analysis methods for multilevel models can be categorized into formula-based methods and simulation-based methods (Murayama, Usami, \& Sakaki, 2022).
Formula-based methods rely on often complicated formulas that can be used to directly calculate power while simulation-based methods rely on repeatedly simulating data with a known true effect size and estimating power empirically (i.e., how often the hypothesis test is significant for the simulated data).
Currently available formula-based software packages for power analysis often do not include GLMMs or are limited to very simple designs (Murayama et al., 2022; Westfall, Kenny, \& Judd, 2014), making it necessary to build data simulations tailored specifically to the study design.
A number of tutorials have been published describing how to perform such simulation-based power analysis for multilevel models (Arend \& Schäfer, 2019; Brysbaert \& Stevens, 2018; DeBruine \& Barr, 2021; Kumle, Võ, \& Draschkow, 2021; Lafit et al., 2021; Zimmer, Henninger, \& Debelak, 2022).
However, most of these tutorials focus on linear mixed models (LMMs) and the most common designs (but see Kumle et al., 2021 for a tutorial that also covers more advanced settings). This narrow focus provides limited guidance for researchers faced with more complex study designs, especially when little prior knowledge about plausible effect sizes is available (see the discussion in Kumle et al., 2021).
The necessary presumptions for simulation-based power analysis with GLMMs include assumptions about the distributional form of the outcome variable, the random effects, and the correlation structure. The distributional assumption specifies the distributional family for the outcome variable (when conditioned on all fixed and random effects). Assumptions about the random effects include the assumption of normality (i.e., that the random effects follow a normal distribution) and the covariance structure among the random effects (i.e., if and how they are correlated). Interpreting these presumptions entails understanding the underlying presumptions of the model and ensuring they align with the characteristics of the data being analyzed.
Existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, which makes specifying assumptions a practical challenge Kumle et al. (2021).

Based on the need for well-powered experimental research using GLMMs and the lack of tools to conduct corresponding power analyses, in this tutorial paper, we present a case study that serves as a practical demonstration of how to perform a simulation-based a priori power analysis with GLMMs. Thereby, we aim to equip researchers with the tools needed to simulate data and determine appropriate sample sizes for their own research.

@@

In this section, we outline the steps for performing data simulation and a priori power analysis for GLMMs using a case study based on a specific experimental study design from the area of human-AI (artificial intelligence) interaction research. The corresponding code in this manuscript is available in the project's repository on the Open Science Framework (\url{https://osf.io/dhwf4/}).

@@

In the present case study, we simulate data for an experiment where the diagnostic performance of users of an AI-enabled diagnostic decision support system will be evaluated.
The goal is to understand how AI advice influences medical decision-making. Participants, radiologists (task experts) and students/interns (non-task experts), review head computer tomography (CT) scans to assess the presence of a bleeding.
To support their decision-making, an AI model provides initial diagnostic advice, which can be used as guidance by the participants. This AI advice can be either correct (80\% of cases) or incorrect (20\%). In the control condition, no AI advice is presented, meaning that the participants have to read the CT scan without any support. After reviewing the CT scan, participants deliver a medical diagnosis (bleeding or no bleeding), which may be either accurate or inaccurate.
This experimental design introduces some missing values by design since the advice is neither correct nor incorrect when no advice is present, which must be taken into account when simulating and analyzing the data. With this experiment, we want to determine if (a) experts are better than non-experts in reading head CT scans and if (b) correct AI advice leads to better diagnostic accuracy than incorrect AI advice.
In this example, recruiting task experts (i.e., radiologists) is more challenging due to their limited availability, while non-experts (i.e., students/interns) are more readily accessible.
The goal of the present simulation-based power analysis is to determine how many task experts and non-experts must be recruited to achieve sufficient statistical power in the planned experiment.

@@

In our case study, we use the lme4 R package (Bates, Mächler, Bolker, \& Walker, 2015), which is a state-of-the-art tool for fitting frequentist GLMMs.\footnote{For Bayesian GLMMs, the brms R package is currently the most prominent option (Bürkner, 2017).}
The lme4 package includes a function called \texttt{simulate} that allows researchers to simulate the dependent variable based on the same model formula used for model fitting, enabling simulation-based power analyses and other related analyses.

However, the model parameterization used by the lme4 package is quite technical, making it difficult for applied researchers to determine whether their specified population model (i.e., the theoretical model that describes the underlying data generation process for a specific population of interest) implies plausible associations in their simulated data. Therefore, in this tutorial, we simulate data for GLMMs from first principles (i.e., creating synthetic data step by step instead of using black box functions) to assist applied researchers in better understanding all model assumptions and then use lme4 to analyze the simulated datasets.\footnote{A less flexible alternative would be to use the simr package (Green \& MacLeod, 2016), which can be used to both simulate data and perform power analysis for models supported by the lme4 package.}

@@

In a GLMM, the expected value of the dependent variable Y conditioned on the vector of predictor variables \(\mathbf{X}\) and random effects \(\mathbf{U}\), transformed by a link function \(g()\) is modeled as a linear combination \(\eta\) of the predictor variables \(\mathbf{X}\), the random effects \(\mathbf{U}\) and the model parameters \(\mathbf{\beta}\) (Fahrmeir et al., 2021): \[
g(E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = \eta
\] Equivalently, the conditional expected value is modeled as the linear combination \(\eta\), transformed by the inverse link function \(g^{-1}()\): \[
E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = g^{-1}(\eta)
\] If the dependent variable (i.e., diagnostic decision) \(Y\) is a binary variable with values \(0\) (i.e., inaccurate), or \(1\) (i.e., accurate), the conditional expected value is equivalent to the probability: \[
P_{si} := P(Y = 1|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})
\] In our case study, \(P_{si}\) is the conditional probability that a subject \(s\) gives the correct response to item (i.e., CT scan) \(i\).

@@

In our case study, the probability of making an accurate diagnostic decision is assumed to depend on the predictors:

\begin{itemize}
\tightlist
\item
  \(advice\_present_{si}\): whether subject \(s\) was presented with AI advice (1) or not (0) when asked to assess item \(i\)
\item
  \(advice\_correct_{si}\): whether this advice was correct (1) or not (0)
\item
  \(expert_s\): whether subject \(s\) was a task expert (1) or not (0)
\end{itemize}

@@

In the GLMM literature, this would be called a binomial GLMM with two random intercepts (for subjects and items), two level-1 predictors (\(advice\_present\), \(advice\_correct\)), one level-2 predictor (\(expert\)) and two cross-level interactions (\(expert \cdot advice\_present\), \(expert \cdot advice\_correct\)).
To limit complexity, we do not consider random slopes, additional predictors or higher-level interactions.

@@

In the first six lines of the function definition, we set some default parameter values (which we will explain in a later section) and load the packages we use to manipulate and simulate data.
In our case study, each subject (\texttt{n\_subjects} in total) is assumed to respond to each item (i.e., CT scan; \texttt{n\_items} in total). Thus, the \texttt{add\_random} command creates a fully-crossed \texttt{data.frame} with \texttt{n\_subjects} \(\times\) \texttt{n\_items} rows. We add a between-subject effect with the \texttt{add\_between} command, simulating that about \(25\%\) of subjects are experts. The next two lines simulate that in \(\frac{2}{3}\) of trials, subjects will be presented with AI advice, and if advice is presented, the advice will be correct in about \(80\%\) of cases (the variable \texttt{advice\_correct} is always 0 when no advice is presented). Next, we simulate one random effect for each subject (\texttt{u0s}) and for each item (\texttt{u0i}). As assumed by standard GLMMs, the \texttt{add\_ranef} function draws the random effects from a normal distribution with a mean 0 and a standard deviation specified by the user. With all design variables done, we are ready to simulate our model equation outlined in the last section. The linear predictor variable \texttt{linpred} (\(\eta\) in the GLMM model equations) combines the predictor variables, random effects, and model parameters as assumed by our model. We then transform the linear predictor with the inverse-link function to compute \texttt{y\_prob}, the probability that the subject correctly solved the item (in R, the inverse-logit link is computed with \texttt{plogis} and the logit link with \texttt{qlogis}). In the final step, we simulate the binary dependent variable \texttt{y\_bin} (i.e., whether the subject makes an accurate diagnostic decision for the CT scan) by -- for each trial -- drawing from a Bernoulli distribution with success probability \texttt{y\_prob}.

@@

In this section, we show how to fit a GLMM with lme4, interpret the model, and test hypotheses derived from a research question.
We simulate data according to our model, in which 100 subjects respond to 50 items (we use \texttt{set.seed} to make the simulation reproducible). However, for the sake of the exercise, we can imagine that this would be real data resulting from our future experiment and think about how we would analyze this data.

@@

\begin{verbatim}
## Generalized linear mixed model fit by maximum likelihood (Laplace
##   Approximation) [glmerMod]
##  Family: binomial  ( logit )
## Formula: 
## y_bin ~ 1 + expert + advice_present + advice_correct + expert:advice_present +  
##     expert:advice_correct + (1 | subject) + (1 | item)
##    Data: dat
## 
##      AIC      BIC   logLik deviance df.resid 
##   4149.4   4201.6  -2066.7   4133.4     4992 
## 
## Scaled residuals: 
##     Min      1Q  Median      3Q     Max 
## -5.7669  0.2125  0.3046  0.4317  2.1056 
## 
## Random effects:
##  Groups  Name        Variance Std.Dev.
##  subject (Intercept) 0.3148   0.5611  
##  item    (Intercept) 0.1624   0.4029  
## Number of obs: 5000, groups:  subject, 100; item, 50
## 
## Fixed effects:
##                       Estimate Std. Error z value Pr(>|z|)    
## (Intercept)             1.0339     0.1103   9.374  < 2e-16 ***
## expert                  1.1849     0.2096   5.654 1.57e-08 ***
## advice_present         -1.3436     0.1206 -11.143  < 2e-16 ***
## advice_correct          2.6154     0.1273  20.540  < 2e-16 ***
## expert:advice_present   1.0589     0.2940   3.601 0.000317 ***
## expert:advice_correct  -1.8104     0.2915  -6.211 5.27e-10 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Correlation of Fixed Effects:
##             (Intr) expert advc_p advc_c exprt:dvc_p
## expert      -0.377                                 
## advic_prsnt -0.349  0.176                          
## advic_crrct  0.023  0.001 -0.668                   
## exprt:dvc_p  0.143 -0.448 -0.412  0.276            
## exprt:dvc_c -0.008  0.004  0.292 -0.435 -0.686
\end{verbatim}

@@

\begin{verbatim}
## [1] 0.95
\end{verbatim}
